---
title: "Access and analysis of ISTAC data through the use of R and Shiny"
author: "Christian González-Martel <br/>José M. Cazorla-Artiles<br/>Carlos J. Perez-González"
date: "2018/07/08"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


```{r xaringan-themer, include = FALSE}
# https://github.com/gadenbuie/xaringanthemer/blob/master/vignettes/css-help.Rmd
library(xaringanthemer)
mono_accent(
  base_color = "#43418A",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono")
)
```


class: inverse, center, middle

# WHY DO THIS PROJECT?

---

# Introduction

The main objective in this work is to show an alternative to create open data portals in public administration.
So that, it can ensure that certain data and information are accessible and available to everyone in an interactive way.

--

As example case, the datasets provided by Canary Institute of Statistics ([ISTAC](http://www.gobiernodecanarias.org/istac/)) are taken.

--

Two steps are considered:

--

- [istacr](https://github.com/rOpenSpain/istacr): R package to make programatic access to data from the  ([ISTAC](http://www.gobiernodecanarias.org/istac/)) base API.

--

- [Canary Islands Tourism Dashboard](https://jmcartiles.shinyapps.io/canary_islands_tourism_dashboard/): A Shiny dashboard to disseminate data and insights from the [Touristic Expenditure Survey of the ISTAC](http://www.gobiernodecanarias.org/istac/temas_estadisticos/sectorservicios/hosteleriayturismo/demanda/C00028A.html).


---

# Motivation

The seed of this project comes from New Zealand. In precise it comes from the [New Zealand Tourism Dashboard](http://tourismdashboard.mbie.govt.nz/), full code available on [github](https://github.com/nz-mbie/tourism-dashboard-public), a really awesome project.

--

Actually, tourism is the main sector in the economic development in Canary Islands, representing the [34.3% GDP and 39.7% employment in this region in 2016 (IMPACTUR, 2016)](http://www.exceltur.org/wp-content/uploads/2017/09/IMPACTUR-Canarias-2016.pdf).

This means that it is necessary to increase the knowledgement of this sector and disseminate information obtained with population in general and decision-makers in concrete.

--

A combination of, by one side open data provided by [ISTAC](http://www.gobiernodecanarias.org/istac/), and by the other side, different tools and programming languages such as [Github](https://github.com/) and [R](https://www.r-project.org/), let us to concrete these ideas in a tourism dashboard with open data, open code and free use.

---

class: inverse, middle, center

# STEP 1: DATA ACCESS <br> The istacr package

---

Starting point consist in create [istacr](https://github.com/rOpenSpain/istacr) package to make programatic access to all datasets available using ISTAC base API. Actually this package is a contribution to the open data community [rOpenSpain](https://ropenspain.es/).

# How to use istacr?<br>Quite easy...

Here we reproduce a [guide of use](http://jmcartiles.netlify.com/post/istacr-a-quick-example-i/). The example of use is focused on showing the total tourism expenditure in Canary Islands by year and by origin country.

---

# Installing the package

First it is needed to download the package from Github and load the needed packages for this example, it means, *dplyr* and *ggplot2*.


```{r comment='#'}
# install and load istacr
# install.packages("devtools")
# library(devtools)
# install_github("rOpenSpain/istacr")
library(istacr)

# Load essentials package for this example
# install.packages("dplyr")
# install.packages("ggplot2")
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```


---

# Using metadata to choose the desired dataset (1)

In this package it is included a dataset which contain some metadata of each dataset available using the [ISTAC](http://www.gobiernodecanarias.org/istac/) base API. It is shown using:

```{r comment="#"}
glimpse(cache)
```


---


# Using metadata to choose the desired dataset (2)

--

As in this example the objective is to get data of tourism expenditure. It is a chance to look for the string *“egt”* in the metadata and generate a subset with all datasets where that string is found. In this case the field (variable) should be *“datos publicadosII”*.

```{r}
busqueda.egt <- istac_search("egt", fields = "datos publicadosII")
# View(busqueda.egt) # use this to see metadata obtained
```

--

When a subset of metadata is get, it is the momment to see the desired *“ID”*, in our case it is:

```{r}
busqueda.egt$ID[1]
```

---

# Downloading data from [ISTAC](http://www.gobiernodecanarias.org/istac/)

--

In one line the full dataset it is ready to use:

```{r}
df <- suppressWarnings(istac(busqueda.egt$ID[1],
                             POSIXct = TRUE, freq = "anual"))
```

--

The next step to show total expenditure, data must be filter the data as follow:

```{r}
df.anual <- filter(df,
                   nchar(Periodos) == 4 &
                     Indicadores == "Valor absoluto" &
                     `Indicadores de gasto` == "GASTO TOTAL" &
                     `Países de residencia` != "TOTAL PAÍSES")
```


---

# Plot data using ggplot

--

Finally, using *ggplot* the year evolution of total expenditure is:

```{r, eval=FALSE, echo=TRUE}
ggplot(df.anual, aes(x = Periodos, y = valor,
                     fill = `Países de residencia`)) +
  geom_col(position = "dodge") +
  xlab("") +
  ylab("") +
  theme(
    panel.background = element_blank(),
    axis.ticks = element_blank()
  )
```

---

class: center

```{r, echo=FALSE}
ggplot(df.anual, aes(x = Periodos, y = valor,
                     fill = `Países de residencia`)) +
  geom_col(position = "dodge") +
  xlab("") +
  ylab("") +
  theme(
    panel.background = element_blank(),
    axis.ticks = element_blank()
  )
```

---

class: inverse, middle, center

# STEP 2: DATA VISUALIZATION
# The Canary Islands Tourism Dashboard

---

# Creating a Shiny Web Application

This application is feed up using the ISTAC base API through istacr and its main purposes are the access to ISTAC tourism data and to facilitate the understanding of tourism patterns in the Canary Islands with special attention to exploit the statistics which comes from Touristic Expenditure Survey.

--

The structure of this Shiny shows a dashboard with main results and several slides describing different data. It is available to filter data and change the rendered visualization. It has an option to download data.

---

class: inverse, middle, center

# THANKS!

---





